---
title: "Week 3 Materials"
format:
  html: default
---
```{r setup, include=FALSE}
# List of packages to install
install.packages("tidyverse")
```
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig.width: 12
#| fig.height: 4

library(tidyverse)
library(igraph)
library(ggraph)
library(patchwork)

# Function to generate network data with a central node
generate_network <- function(n_nodes, n_edges, seed = 42) {
  set.seed(seed)
  nodes <- tibble(id = 0:n_nodes)  # 0 is the central node (the student)
  edges <- tibble(
    from = c(rep(0, n_nodes), sample(1:n_nodes, n_edges - n_nodes, replace = TRUE)),
    to = c(1:n_nodes, sample(1:n_nodes, n_edges - n_nodes, replace = TRUE))
  ) %>% filter(from != to) %>% distinct()
  
  graph <- graph_from_data_frame(edges, directed = FALSE, vertices = nodes)
  
  # Calculate closeness centrality
  closeness <- closeness(graph, normalized = TRUE)
  V(graph)$closeness <- closeness
  V(graph)$is_central <- as.numeric(V(graph)$name == "0")
  
  return(graph)
}

# Function to plot a network
plot_network <- function(graph, title) {
  ggraph(graph, layout = "fr") +
    geom_edge_link(alpha = 0.4) +
    geom_node_point(aes(size = closeness, color = as.factor(is_central)), alpha = 0.7) +
    scale_size_continuous(range = c(3, 10)) +
    scale_color_manual(values = c("0" = "darkblue", "1" = "red")) +
    theme_graph() +
    labs(title = title) +
    theme(plot.title = element_text(hjust = 0.5),
          legend.position = "none")
}

# Generate three different networks
slim_network <- generate_network(5, 8, seed = 1)
frantic_network <- generate_network(20, 50, seed = 2)
healthy_network <- generate_network(15, 25, seed = 3)

# Create individual plots
p1 <- plot_network(slim_network, "Slim")
p2 <- plot_network(frantic_network, "Frantic")
p3 <- plot_network(healthy_network, "Healthy")

# Combine plots
combined_plot <- p1 + p2 + p3 +
  plot_layout(ncol = 3) +
  plot_annotation(
    title = "Social Connectedness Networks of Three Students",
    subtitle = "Node size represents closeness centrality. Red node is the student.",
    theme = theme(plot.title = element_text(hjust = 0.5),
                  plot.subtitle = element_text(hjust = 0.5))
  )

# Display the combined plot
combined_plot
```

## Overview

This week we'll be covering mostly big picture background content about
the technical side of things. We'll begin with background on data
science generally then narrow to talking about why this course is taught
in R (rather than some other coding language). We will touch on how
Excel will likely remain a valuable tool to teach yourself, and then
I'll give you some hard-won tips on problem solving so that you can
begin to develop strategies for handling the errors and warnings that
you'll inevitably encounter as you grow your competency in this arena.
In lab, we'll discuss some crucial components of navigating and using
RStudio as well as a very gentle first foray into it's use.

## Lecture 3 - Data Visualisation

<iframe class="slide-deck" src="test.html" height="420" width="750" style="border: 1px solid #2e3846;">

</iframe>

[Lecture Slides -- Full Screen](test.qmd)

## Lab 03 - Excel, RStudio & ggplot2

### Learning Objectives 

Upon completion of this lab, students will be able to:

**Evaluate** your Excel proficiency and its relevance to potential
career paths

**Calculate** basic descriptive statistics (mean, count) using Excel

**Differentiate** between levels of measurement and types of data in
research methods

**Explain** measures of central tendency and variability

**Familiarise** yourself with RStudio and perform basic operations

<iframe class="slide-deck" src="slides_3b.html" height="420" width="750" style="border: 1px solid #2e3846;">

</iframe>

[Lab Slides -- Full Screen](slides_3b.qmd)

## DataSkills 03

Learning Objectives:

**Start Literate Coding** by blending narrative and code chunks

**Copy** a simple document in R, incorporating data analysis and
commentary

**Render** documents into HTML or PDF formats

**Recognize** the role of tools like R and RStudio in Open Science
practices
