---
title: "Intro to Data Visualisation II"
author: "Lecture 3 -- More than Graphs"
engine: knitr

format: 
  revealjs: 
    slide-number: c/t
    controls-layout: bottom-right
    embed-resources: true
    scrollable: true
    code-overflow: wrap
    code-line-numbers: true
    code-copy: true
    theme: [simple, slides.scss]
    reference-location: document
    footer: "[Data Visualisation II](https:www.bbc.co.uk)"
webr:
  packages: ['ggplot2', 'dplyr'] # Install R packages on document open
# autoload-packages: false       # Disable automatic loading of packages
# show-startup-message: false    # Disable displaying status of webR initialization
filters:
- webr
---

---
title: "Describing Data"
format: revealjs
---

## Introduction

> "Far better an approximate answer to the right question, which is often vague, than an exact answer to the wrong question, which can always be made precise." ---John W. Tukey

- Descriptive statistics are tools for describing data
- There are many ways to describe data
- Choose the most useful way for your data
- Describing data is necessary because there's usually too much of it

---

## Too Many Numbers

- Example: Asking thousands of people about their happiness
- Raw data is overwhelming and difficult to interpret

```{r}
#| echo: false
happiness <- rnorm(500, 100, 500)
knitr::kable(matrix(round(happiness), ncol=10, nrow=5))
```

---

## Looking at Data: Graphs

### Scatter Plot

```{r}
#| echo: false
plot(happiness)
```

---

## Looking at Data: Histograms

```{r}
#| echo: false
hist(happiness)
```

- Shows the distribution of data
- Reveals shape, center, and spread

---

## Important Ideas

1. Distribution
2. Central Tendency
3. Variance

---

## Measures of Central Tendency

1. Mode
2. Median
3. Mean

---

## Mode

- Most frequently occurring number
- Example: 1, 1, 1, 2, 3, 4, 5, 6
- Mode = 1

---

## Median

- Middle value when data is ordered
- Example: 1, 3, 4, **5**, 6, 7, 9
- Median = 5

---

## Mean

$Mean = \bar{X} = \frac{\sum_{i=1}^{n} x_{i}}{N}$

- Sum of all values divided by the number of values
- Represents the "average"

---

## Measures of Variation

1. Range
2. Variance
3. Standard Deviation

---

## Range

- Minimum and maximum values
- Example: 1, 3, 4, 5, 5, 6, 7, 8, 9, 24
- Range: 1 to 24

---

## Variance

$variance = \frac{\text{Sum of squared difference scores}}{\text{Number of Scores}}$

- Measures spread of data around the mean

---

## Standard Deviation

$\text{standard deviation} = \sqrt{\frac{\sum_{i}^{n}({x_{i}-\bar{x})^2}}{N}}$

- Square root of variance
- In same units as original data

---

## Using Descriptive Statistics

- Reduce large datasets to summary statistics
- Combine with graphical representations
- Be aware of limitations (e.g., Anscombe's Quartet)

---

## Anscombe's Quartet

```{r}
#| echo: false
library(ggplot2)
library(data.table)

ac <- fread("data/anscombe.txt")
ac <- as.data.frame(ac)

ac_long <- data.frame(
  x = c(ac[,1], ac[,3], ac[,5], ac[,7]),
  y = c(ac[,2], ac[,4], ac[,6], ac[,8]),
  quartet = as.factor(rep(1:4, each=11))
)

ggplot(ac_long, aes(x=x, y=y, color=quartet)) +
  geom_point() +
  theme_classic() +
  facet_wrap(~quartet)
```

- Same descriptive statistics, different patterns

---

## Remember

- Always look at your data
- Combine descriptive statistics with visualizations
- Be aware of potential hidden patterns