---
title: "Week 03 Worksheet - part 1"
subtitle: "A first gentle introduction"
author: "INSERT YOUR NAME HERE"
format: 
  html: default
  pdf: default
editor: visual
---

# Welcome to RStudio!

This worksheet will introduce you to the RStudio interface and the
concept of literate programming using R.

## The RStudio Interface (Lab Tutor Tour)

When you open RStudio, you'll see four main panes:

1.  **Source Editor** (top-left): This is where you write and edit your
    code and text.
2.  **Console** (bottom-left): This is where R commands are executed and
    where you see the output.
3.  **Environment/History** (top-right): This shows the objects in your
    workspace and your command history.
4.  **Files/Plots/Packages/Help** (bottom-right): This multi-purpose
    pane shows your files, plots, installed packages, and help
    documentation.

## This is a .qmd file

It is markdown - a text only language that can be edited anywhere, even
on your phone, because it uses commonly occurring symbols to do all the
formatting. If you happen to know html, it's pretty similar. It's simple
(once you get the idea), very small in terms of file-size, easily
accessed, and versatile.

## Understanding Literate Programming

Literate programming is a paradigm that combines explanatory text with
executable code in a single document. This approach, pioneered by Donald
Knuth in 1984, aims to make programming more accessible, understandable,
and maintainable. When you learn about Open Science practices next week
with Stacey, she will probably talk about sharing of data, replicability
and other big issues. One way in which we can make Science more
accessible and open is by using techniques like this.

### Key Concepts

1.  **Integration of Code and Documentation**: In literate programming,
    the code is interspersed with narrative text that explains the
    purpose and functionality of the code.

2.  **Human-Oriented**: The primary focus is on making the program
    understandable to humans, rather than just computers.

3.  **Executable Documents**: The resulting document can be both read as
    a coherent explanation and executed as a functional program.

### Benefits in Data Science and Research

Literate programming is particularly valuable in data science and
research for several reasons:

1.  **Reproducibility**: By combining code, results, and explanations,
    others can easily reproduce and verify your work.

2.  **Clear Communication**: It allows you to explain your thought
    process, methodology, and interpretation of results alongside the
    code that generates them.

3.  **Error Checking**: The close proximity of code and explanation
    makes it easier to spot inconsistencies or errors.

4.  **version Control**: Changes in both code and narrative can be
    tracked together, providing a comprehensive history of the project's
    evolution.

### Example in R

Here's a simple example of literate programming in R using a Quarto
document:

In this analysis, we'll explore the relationship between a car's
horsepower and its fuel efficiency using the `mtcars` dataset (this is a
commonly used dataset that comes installed in R).

## "The Results of My Amazing Experiment"

Dear Reader, be amazed at my wonderfulness! Below, I am going to walk
you through the analysis. If you are reading this as a .qmd file, you
will see all the code. But after I render it to html or pdf, you'll only
see what I want you to see, such as just the pretty pictures, without
the code! Or I can fold up the code here too!

Another aspect of literate programming, other than narrating the larger
document, you can also add comments in the code, to draw an interested
reader to specialist aspects of what's happening there, or just to
signal what the code is doing. When you try to work with someone else's
code, comments like the ones below (but better) are really helpful!

If you are reading along in the .qmd file, press the green arrow to the
right of the code cell to execute it. And bingo! R does it's work and
performs the calculations.

```{r}
# Load the mtcars dataset
data(mtcars)

# Create a scatter plot
plot(mtcars$hp, mtcars$mpg, 
     main="Horsepower vs. Fuel Efficiency", # Clearly these are the most important things!
     xlab="Horsepower", 
     ylab="Miles per Gallon", # Can we change this to Kilometers per Litre for other countries? 
     pch=19, 
     col="darkblue") # The best colour for car-based statistics
```

## The Visual Editor

RStudio's visual editor provides a user-friendly interface for creating
Quarto documents. It allows you to write text and code in a "What You
See Is What You Get" (WYSIWYG) environment.

To switch to the visual editor, click the "Visual" button at the
top-left of the Source Editor.

## Writing Narrative Text

In the visual editor, you can simply type your text as you would in any
word processor. You can use the formatting toolbar at the top to:

-   Apply **bold** or *italic* formatting
-   Create bullet or numbered lists
-   Add headings
-   Insert links or images

Try writing a short paragraph about why you're studying Psychology
below:

\[Your text here\]

## Adding Code Cells

To add a code cell:

1.  Click the "+C" button in the toolbar or use the keyboard shortcut
    Cmd+Option+I (Mac) or Ctrl+Alt+I (Windows/Linux)
2.  You'll see a new code cell appear
3.  Type your R code inside this cell

Let's try a simple calculation. Add a code cell and type the following:

```{r}
2 + 2
```

## Running Code

To run the code in a cell:

Click the "Run" button (green play icon) at the top-right of the cell,
or Use the keyboard shortcut: Cmd+Enter (Mac) or Ctrl+Enter
(Windows/Linux)

The output will appear directly below the code cell. Rendering Your
Document To create the final document:

Click the "Render" button at the top of the editor Choose your desired
output format (HTML or PDF) RStudio will process your document and
display the result

## Exercise

Load the data Gordon has been using today.
<https://github.com/DrDeception/Week0203/blob/3276cb8fa56a466a3e25761523cea475ed4ee4bd/materials/data/Y1W3_data.csv>

```{r}
# Welcome to R! This is a comment. R doesn't run anything on a line after a '#'.

# Before we can read our data, we need a special tool called 'readr'.
# Think of this like buying a book from a bookstore.
# We only need to do this once, just like you only need to buy a book once.
install.packages("readr")

# Now that we've bought the book (installed the package), we need to take it off our shelf to use it.
# In R, we do this by 'loading' the library. We'll need to do this each time we start a new R session.
# It's like taking the book off your shelf each time you want to read it.
library(readr)  # This is like opening the book 'readr' to use its contents

# Now let's specify where our data is located. This is like writing down the address of a library.
url <- "https://raw.githubusercontent.com/DrDeception/Week0203/3276cb8fa56a466a3e25761523cea475ed4ee4bd/materials/data/Y1W3_data.csv"

# We use the read_csv() function from our 'readr' book to load the data.
# This is like using the skills we learned from the book to read a document in the library.
data <- read_csv(url)

# Let's take a quick look at the first few rows of our data.
# This is like skimming the first page of the document.
print(head(data))

# Now, let's get a summary of our data. 
# This is like reading the table of contents and index to get an overview.
print(summary(data))

# We can check how big our dataset is.
# This is like counting the pages and chapters in our document.
cat("Our dataset has", dim(data)[1], "rows (like pages) and", dim(data)[2], "columns (like chapters)\n")

# Finally, let's see what information (columns) we have in our dataset.
# This is like looking at the headings in our document.
cat("Our dataset contains information about:\n")
print(colnames(data))

# Congratulations! You've just used R to bring in data from the internet and take a first look at it.
# It's like you've gone to a digital library, found a document, and skimmed through its contents!
```

## Normally... 

That code chunk would only be like this:

```{r}
install.packages("readr")
library(readr)
data <- read_csv("https://raw.githubusercontent.com/DrDeception/Week0203/3276cb8fa56a466a3e25761523cea475ed4ee4bd/materials/data/Y1W3_data.csv")
summary(data) # Even this isn't necessary if you know what the data includes already!
dim(data) # or this
colnames(data) # or this.
```

## Data Viewer

If you look in the Environment panel to the right, you will see `data`
and if you click on the little spreadsheet icon to the right, you can
look at it like a spreadsheet!

### My First Plot

```{r}
# COPY SOME CODE IN HERE from the slides from this morning - choose one of the early, simpler plots if possible
```

## Run the code cell

Render your document to html (website) or pdf (document) and view the
results!

## Conclusion

You've now learned the basics of:

::: callout-note
-   Navigating the RStudio interface Using the visual editor for Quarto
    documents
-   Writing narrative text and adding formatted elements
-   Inserting and running code cells to do sums and create graphs
-   Rendering your document
:::

Keep practicing these skills as you continue your journey with R and
data analysis!
